Analysis of Temporal Trends using Bayesian Generalized Additive Models

### 2025-04-29 analysis/code/010-summary-of-gams.qmd
nearly got all the gams diagnostics done

Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @R_core]. `brms` utilizes Stan (version `r brms::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].

@@

results_brms <- over_time_long_colour_gams_testing_nested_df %>%
  mutate(
    # Create a new column for brms models to avoid overwriting gam models if needed
    mod_brms = map2(data, variable, ~{
      current_df <- .x        # The dataframe for the current row
      current_var <- .y       # The variable name for the current row
      response_col <- "value" #  this matches the response column name in nested dfs

      message(paste("Fitting brms model for:", current_var)) # Progress message

      # --- 1. Choose the brms family based on the variable name ---
      chosen_family <- if (current_var == "N. of authors") {
        negbinomial() # For count data
      } else if (current_var == "N. of pages") {
        negbinomial() # For count data
      } else if (current_var == "Recency of references") {
        zero_one_inflated_beta() # For 0/1 inflated proportions
      } else if (current_var == "Diversity of references") {
        gaussian() # For continuous, approx normal data
      } else if (current_var == "Relative title length (ln)") {
        gaussian() # For continuous, normal data
      } else {
        warning(paste("Unknown variable:", current_var, "- using Gaussian default."))
        gaussian() # Default fallback
      }

      # --- 2. Construct the formula using brms::bf() ---
      # Handle the special structure for Zero-One Inflated Beta
      if (inherits(chosen_family, "zero_one_inflated_beta")) {
        # Formula for the mean (between 0 and 1), plus formulas for
        # zoi (zero-inflation probability) and coi (one-inflation probability).
        # zoi and coi are modelled simply as intercepts.
        brm_formula <- bf(
          paste(response_col, "~ s(year, bs = 'cr')"), 
          # Smooth term for mean
          zoi ~ 1,                                    
          # Intercept for prob of 0
          coi ~ 1                                     
          # Intercept for prob of 1
        )
      } else {
        # Standard formula for other families
        brm_formula <- bf(
          paste(response_col, "~ s(year, bs = 'cr')")
        )
      }

      # --- 3. Fit the brms model ---
      # Using tryCatch to prevent one failure from stopping everything
      tryCatch({
        brm(
          formula = brm_formula,
          family = chosen_family,
          data = current_df,
          # --- Stan control arguments (adjust as needed) ---
          cores = 4,  # Number of cores for parallel chains 
          chains = 4, # Number of Markov chains (4 is standard for final runs)
          iter = 3000, # Total iterations per chain (includes warmup)
          warmup = 1000,# Number of warmup iterations (discarded)
          control = list(adapt_delta = 0.9999), # for helping convergence for complex models
          seed = 123, # For reproducibility
          silent = 2, # Suppress compilation messages from Stan
          refresh = 0 # Suppress iteration progress updates

        )
      }, error = function(e) {
        # Print a warning if a model fails to fit
        warning(paste("brms fitting failed for variable:", 
                      current_var, "\nError:", e$message))
        return(NULL) # Return NULL for failed models
      })
    }) # End of map2
  ) # End of mutate

@@

# Load my brms results object so we don't have to re-run the above block
load(here::here("analysis/data/results_brms.RData"))

# For demonstration, let's assume 'results_brms' exists and contains valid models
# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))

@@

Models were fitted using `r results_brms_valid$mod_brms[[1]]$algorithm$chains` Markov chains. Each chain consisted of `r results_brms_valid$mod_brms[[1]]$algorithm$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$algorithm$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$algorithm$iter - results_brms_valid$mod_brms[[1]]$algorithm$warmup) * results_brms_valid$mod_brms[[1]]$algorithm$chains` posterior samples per model for inference. No thinning was applied (thin = 1).

@@

1.  **Divergent Transitions:** All final models reported **zero divergent transitions** after warmup, indicating the sampler could explore the posterior distribution without significant issues. Initial fits for 'Recency of references' showed divergent transitions, which were resolved by refitting with `control = list(adapt_delta = 0.995)`.")

```{r check_divergences, echo=TRUE}
#| tbl-cap: "Number of divergences for each variable"

# Code to programmatically check divergences 
library(rstan)
sampler_params_list <- map(results_brms_valid$mod_brms, ~{
  if (!is.null(.x) && inherits(.x, "brmsfit")) {
    # Get parameters excluding warmup draws
    nuts_params(.x, inc_warmup = FALSE)
  } else {
    NULL
  }
})

divergences_per_model <- map_dbl(sampler_params_list, ~{
  if (!is.null(.x)) {
    # Sum divergent transitions across all chains
    sum(subset(.x, Parameter == "divergent__")$Value)
  } else {
    NA_real_ # Use NA for models that didn't fit
  }
})

divergence_summary <- tibble(
   variable = results_brms_valid$variable,
   n_divergences = divergences_per_model
)

kable(divergence_summary)

@@

#```{r convergence_summary_table, echo=FALSE}
# Load necessary libraries if not already done
library(brms)
library(posterior) # Explicitly load posterior for ess functions
library(purrr)
library(dplyr)
library(knitr)
library(kableExtra) # Optional for styling

# --- Calculate Convergence Diagnostics ---
# Use map2_dfr to iterate over models and variable names, creating a data frame
convergence_summary_df <- map2_dfr(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    # Assign model and variable name for clarity within the loop
    model <- .x
    var_name <- .y

    # Initialize values
    max_rhat <- NA_real_
    min_bulk_ess <- NA_real_
    min_tail_ess <- NA_real_
    n_divergences <- NA_integer_

    # Proceed only if model is a valid brmsfit object
    if (!is.null(model) && inherits(model, "brmsfit")) {
      # Use try() to gracefully handle potential errors
      try({
        # R-hat
        rhats <- brms::rhat(model)
        if(is.numeric(rhats)) max_rhat <- max(rhats, na.rm = TRUE)

        # Bulk ESS - Use posterior::ess_bulk()
        bulk_ess <- posterior::ess_bulk(model)
        if(is.numeric(bulk_ess) && length(bulk_ess) > 0) min_bulk_ess <- min(bulk_ess, na.rm = TRUE)

        # Tail ESS - Use posterior::ess_tail()
        tail_ess <- posterior::ess_tail(model)
         if(is.numeric(tail_ess) && length(tail_ess) > 0) min_tail_ess <- min(tail_ess, na.rm = TRUE)

        # Divergences
        nuts_p <- brms::nuts_params(model)
        nuts_df <- as.data.frame(nuts_p)
        if ("divergent__" %in% names(nuts_df)) {
           n_divergences <- sum(nuts_df$divergent__, na.rm = TRUE)
        } else {
           warning(paste("Column 'divergent__' not found in nuts_params output for:", var_name))
           n_divergences <- 0L
        }
      }, silent = TRUE) # Keep try silent, but check warnings/NA values later
    }

    # Return a tibble
    tibble(
      Variable = var_name,
      `Max R-hat` = max_rhat,
      `Min Bulk ESS` = floor(min_bulk_ess), # Floor ESS for cleaner table
      `Min Tail ESS` = floor(min_tail_ess), # Floor ESS for cleaner table
      `N Divergences` = as.integer(n_divergences) # Ensure integer
    )
  }
)

# --- Display the Table ---
# (kable formatting code remains the same)
kable(
  convergence_summary_df,
  caption = ,
  digits = c(NA, 3, 0, 0, 0), # Specify digits for numeric columns
  booktabs = TRUE
) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

# somthing not quite right here

2.  **R-hat (R̂):** The potential scale reduction factor (R̂) was examined for all parameters. All R̂ values were ≤ 1.01, indicating successful convergence of the chains to a common distribution.
3.  **Effective Sample Size (ESS):** Both Bulk-ESS and Tail-ESS were assessed. Minimum ESS values across all parameters were deemed sufficiently large (e.g., > 1000 *[adjust threshold based on your results/field standards]*) for reliable estimation of posterior summaries and credible intervals.
4.  **Trace Plots:** Trace plots for key parameters were visually inspected and showed good mixing and stationarity, providing further confidence in chain convergence. *(Optionally include an example trace plot in an Appendix)*.

@@

Model fit was primarily assessed using graphical posterior predictive checks (PPCs). Datasets were simulated from the posterior predictive distribution of each fitted model and compared to the observed data distribution.

```{r ppc_plot, fig.cap="Posterior predictive checks (density overlays) for each model. The dark line (y) represents the density of the observed data, while the light lines (y_rep) represent densities from datasets simulated from the fitted model."}
# Generate PPC plots (density overlay is often informative)
ppc_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    model <- .x
    var_name <- .y
    # Use tryCatch in case pp_check fails for some reason
    tryCatch({
      pp_check(model, type = "dens_overlay", ndraws = 100) + # ndraws controls number of light lines
        ggtitle(var_name) +
        theme(plot.title = element_text(hjust = 0.5, size = 10),
              legend.position = "right") # Adjust legend
    }, error = function(e) {
      warning("pp_check failed for: ", var_name, " Error: ", e$message)
      return(NULL)
    })
  }
)

@@

The estimated non-linear effect of publication year on each response variable is visualized below, showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).

```{r smooth_plots, fig.cap="Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs. The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval. Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."}
# Generate conditional smooth plots
conditional_smooths_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    model <- .x
    var_name <- .y
    tryCatch({
      # Generate plot data first
      cs_data <- conditional_smooths(model, plot = FALSE)[[1]] # Get data for the first smooth
      # Create ggplot manually for more control if needed, or use plot() method
      plot_object_list <- plot(conditional_smooths(model), plot = TRUE)

      if (length(plot_object_list) >= 1 && inherits(plot_object_list[[1]], "ggplot")) {
         # Add title and potentially customize axes
         y_axis_label <- case_when(
             var_name %in% c("N. of authors", "N. of pages") ~ "s(year) Contribution (log scale)",
             var_name == "Recency of references" ~ "s(year) Contribution (logit scale)",
             TRUE ~ "s(year) Contribution" # Gaussian models
         )
         p <- plot_object_list[[1]] +
           ggtitle(var_name) +
           ylab(y_axis_label) +
           theme(plot.title = element_text(hjust = 0.5, size = 10))
         return(p)
      } else {
         warning(paste("Could not extract ggplot object for:", var_name))
         return(NULL)
      }
    }, error = function(e) {
       warning("Conditional smooths plot failed for: ", var_name, " Error: ", e$message)
       return(NULL)
    })
  }
)

# Remove NULLs and combine
valid_smooth_plots <- conditional_smooths_plots_list[!sapply(conditional_smooths_plots_list, is.null)]
plot_grid(plotlist = valid_smooth_plots, ncol = 3)

@@

```{r param_summary, echo=FALSE}
# Extract tidy summaries for relevant parameters
tidy_summaries <- map_dfr(
  results_brms_valid$mod_brms,
  ~ tidy(.x, effects = c("fixed", "ran_pars"), conf.int = TRUE, conf.level = 0.95),
  .id = "model_index" # Add an index to link back to variable name
)

# Join with variable names
tidy_summaries <- tidy_summaries %>%
  mutate(model_index = as.integer(model_index)) %>%
  left_join(results_brms_valid %>% select(variable) %>% mutate(model_index = row_number()), by = "model_index") %>%
  select(variable, everything(), -model_index) %>%
  # Optional: Filter specific terms if the table gets too long
  filter(component != "zi") %>% # Often not directly estimated for NB/ZOIB in tidy
  # Rename for clarity
  rename(
    Parameter = term,
    Estimate = estimate,
    Est.Error = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  select(variable, Parameter, Estimate, Est.Error, `Lower 95% CI`, `Upper 95% CI`, component, group)

# Display using kable
kable(
  tidy_summaries,
  caption = "Posterior summary statistics for key model parameters. 'fixed' component includes intercepts, 'ran_pars' includes smooth term standard deviations (sds), 'dist' includes distributional parameters (sigma, shape, phi, zoi, coi).",
  digits = 2,
  booktabs = TRUE
) %>%
 kableExtra::kable_styling(latex_options = c("striped", "scale_down"))

>>>>

Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @R_core]. `brms` utilizes Stan (version `r rstan::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].

@@

# --- Define the Prior ---
# Apply a regularizing prior to the standard deviation of all smooth terms
# Exponential(1) encourages smaller standard deviations (less wiggly smooths)
sds_prior <- prior(exponential(1), class = "sds")

# --- Run the models with the added prior ---
results_brms <- over_time_long_colour_gams_testing_nested_df %>%
  mutate(
    # Create a new column for brms models
    mod_brms = map2(data, variable, ~{
      current_df <- .x        # The dataframe for the current row
      current_var <- .y       # The variable name for the current row
      response_col <- "value" # Ensure this matches

      message(paste("Fitting brms model for:", current_var))

      # --- 1. Choose the brms family ---
      chosen_family <- if (current_var == "N. of authors") {
        negbinomial()
      } else if (current_var == "N. of pages") {
        negbinomial()
      } else if (current_var == "Recency of references") {
        zero_one_inflated_beta()
      } else if (current_var == "Diversity of references") {
        gaussian()
      } else if (current_var == "Relative title length (ln)") {
        gaussian()
      } else {
        warning(paste("Unknown variable:", current_var, "- using Gaussian default."))
        gaussian()
      }

      # --- 2. Construct the formula ---
      if (inherits(chosen_family, "zero_one_inflated_beta")) {
        brm_formula <- bf(
          paste(response_col, "~ s(year, bs = 'cr')"),
          zoi ~ 1,
          coi ~ 1
        )
      } else {
        brm_formula <- bf(
          paste(response_col, "~ s(year, bs = 'cr')")
        )
      }

      # --- 3. Fit the brms model with the prior ---
      tryCatch({
        brm(
          formula = brm_formula,
          family = chosen_family,
          data = current_df,
          prior = sds_prior, 
          cores =    4,
          chains =   4,
          iter =     50000,
          warmup =   5000,
          control =  list(adapt_delta = 0.99999), 
          seed =     123,
          silent =   2,
          refresh =  0
        )
      }, error = function(e) {
        warning(paste("brms fitting failed for variable:",
                      current_var, "\nError:", e$message))
        return(NULL)
      })
    }) # End of map2
  ) # End of mutate

@@

# Load my brms results object so we don't have to re-run the above block
results_brms <- readRDS(here::here("analysis/data/results_brms.RData"))

# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))
```

@@

Models were fitted using `r brms::nchains(results_brms_valid$mod_brms[[1]])` Markov chains. Each chain consisted of `r results_brms_valid$mod_brms[[1]]$fit@sim$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$fit@sim$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$fit@sim$iter - results_brms_valid$mod_brms[[1]]$fit@sim$warmup) * brms::nchains(results_brms_valid$mod_brms[[1]])` posterior samples per model for inference. No thinning was applied (thin = 1).

@@

@@

# Function to extract diagnostics from a single model
get_diagnostics <- function(m) {
  ms <- summary(m)

  tibble(
    chains = brms::nchains(m),
    rhat_max = max(brms::rhat(m)), 
    bulk_ess_min = min(ms$spec_pars$Bulk_ESS, 
                       na.rm = TRUE),
    tail_ess_min = min(ms$spec_pars$Tail_ESS, 
                       na.rm = TRUE),
    num_divergent = rstan::get_num_divergent(m$fit)
  )
}

# Apply to all models
diagnostics_table <- results_brms_valid %>%
  mutate(diagnostics = map(mod_brms, get_diagnostics)) %>%
  unnest(diagnostics) %>% 
  select(variable, 
         chains,
         rhat_max,
         bulk_ess_min,
         tail_ess_min,
         num_divergent)

# View the table

library(knitr)
library(kableExtra)

kable(
  diagnostics_table,
  caption = "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0. Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0.",
  digits = c(NA, 0, 3, 0, 0, 0), # Chains/ESS/Div as integers, Rhat to 3 decimals
  booktabs = TRUE,
  linesep = ""
) %>%
  kableExtra::kable_styling(latex_options = "scale_down", full_width = FALSE) %>%
  kableExtra::footnote(general = "Ideal values: Max R-hat ≈ 1.0, Min ESS > 1000, N Divergences = 0.", general_title = "Note:", footnote_as_chunk = TRUE)
```

1.  **Divergent Transitions:** All final models reported less than ten **divergent transitions** after warmup, indicating the sampler could explore the posterior distribution without significant issues. Initial fits for 'Recency of references' showed divergent transitions, which were managed by refitting with `control = list(adapt_delta = 0.9999)`.")
2.  **R-hat (R̂):** The potential scale reduction factor (R̂) was examined for all parameters. All R̂ values were ≤ 1.01, indicating successful convergence of the chains to a common distribution.
3.  **Effective Sample Size (ESS):** Both Bulk-ESS and Tail-ESS were assessed. Minimum ESS values across all parameters were deemed sufficiently large (> 1000) for reliable estimation of posterior summaries and credible intervals.
4.  **Trace Plots:** Trace plots for key parameters were visually inspected and showed good mixing and stationarity, providing further confidence in chain convergence (@fig-trace-plots-mcmc-trace).

```{r fig.height=3 * length(results_brms_valid$mod_brms)}
#| label: fig-trace-plots-mcmc-trace,
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: "Trace plots for key parameters (Intercept and smooth term standard deviation) with divergence markers for each model. Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)     # For map2
library(dplyr)     # For tibble etc.
library(posterior) # For as_draws_array
library(bayesplot) # For mcmc_trace and nuts_params
library(ggplot2)   # For ggtitle and theme adjustments
library(cowplot)   # For plot_grid
#---------------------------------------------------------------------

#---------------------------------------------------------------------
# Generate Trace Plots using mcmc_trace for each model
#---------------------------------------------------------------------
trace_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    # Assign model and variable name for clarity
    model <- .x
    var_name <- .y

    # Initialize plot object as NULL
    p <- NULL

    # Proceed only if model is a valid brmsfit object
    if (!is.null(model) && inherits(model, "brmsfit")) {
      try({
        # --- Define parameters to plot ---
        # Use the names confirmed to work: b_Intercept and sds_syear_1
        # Check if these parameters actually exist in the current model
        all_model_pars <- parnames(model)
        pars_to_plot <- character(0) # Start with empty vector

        if ("b_Intercept" %in% all_model_pars) {
          pars_to_plot <- c(pars_to_plot, "b_Intercept")
        } else {
           warning(paste("'b_Intercept' not found for model:", var_name))
        }

        if ("sds_syear_1" %in% all_model_pars) {
           pars_to_plot <- c(pars_to_plot, "sds_syear_1")
        } else {
           # Handle cases where the smooth sds name might differ slightly
           # Example: find any parameter starting with sds_s
           sds_par_alt <- grep("^sds_s", all_model_pars, value = TRUE)
           if(length(sds_par_alt) > 0) {
               pars_to_plot <- c(pars_to_plot, sds_par_alt[1])
               warning(paste("Using alternative sds name:", sds_par_alt[1], "for model:", var_name))
           } else {
               warning(paste("'sds_syear_1' (or similar) not found for model:", var_name))
           }
        }
        # --- End of parameter identification ---

        # Proceed only if we found parameters to plot
        if(length(pars_to_plot) > 0) {
            # 1) Extract draws as an array
            post_array <- posterior::as_draws_array(model)

            # 2) Extract NUTS diagnostics (divergences)
            np <- brms::nuts_params(model)

            # 3) Generate trace plot with divergence ticks
            p <- bayesplot::mcmc_trace(
                    post_array,
                    pars = pars_to_plot, # Use the identified parameters
                    np = np, # Provide NUTS parameters for divergence markers
                    facet_args = list(ncol = 1, strip.position = "left") # Arrange vertically
                  ) +
                 ggtitle(var_name) + # Add variable name as title
                 theme(
                    plot.title = element_text(hjust = 0.5, size = 11), # Center title
                    legend.position = "none" # Often hide legend
                    )
        } else {
            message(paste("No parameters selected for plotting trace for model:", var_name))
        }

      }, silent = TRUE) # Use try silently, check for NULL later
    } else {
        warning(paste("Invalid or NULL model object encountered for variable:", var_name))
    }

    # Return the ggplot object (or NULL if failed or no pars found)
    return(p)
  }
)
#---------------------------------------------------------------------

#---------------------------------------------------------------------
# Combine and Display Plots
#---------------------------------------------------------------------
# Remove any NULL elements from the list
valid_trace_plots <- trace_plots_list[!sapply(trace_plots_list, is.null)]

# Check if there are any valid plots left
if (length(valid_trace_plots) > 0) {
  # Combine the valid plots into a grid
  # Adjust ncol and fig.height chunk option as needed
  combined_trace_plots <- plot_grid(
                              plotlist = valid_trace_plots,
                              ncol = 2 # Arrange in 2 columns
                              )

  # Display the combined plot
  print(combined_trace_plots)
} else {
  message("No valid trace plots were generated.")
}
#---------------------------------------------------------------------
```

@@

Model fit was primarily assessed using graphical posterior predictive checks (PPCs). Datasets were simulated from the posterior predictive distribution of each fitted model and compared to the observed data distribution (@fig-ppc-plot).

```{r}
#| label: fig-ppc-plot
#| fig-cap: "Posterior predictive checks (density overlays) for each model. The dark line (y) represents the density of the observed data, while the light lines (y_rep) represent densities from datasets simulated from the fitted model."
#| echo: true  
#| warning: false  
#| message: false  
# Generate PPC plots (density overlay is often informative)
ppc_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    model <- .x
    var_name <- .y
    # Use tryCatch in case pp_check fails for some reason
    tryCatch({
      pp_check(model, type = "dens_overlay", ndraws = 100) + # ndraws controls number of light lines
        ggtitle(var_name) +
        theme(plot.title = element_text(hjust = 0.5, size = 10),
              legend.position = "right") # Adjust legend
    }, error = function(e) {
      warning("pp_check failed for: ", var_name, " Error: ", e$message)
      return(NULL)
    })
  }
)

@@

@@

```{r}
#| label: tab:param-summary
#| echo: false
#| message: false
#| warning: false
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)
library(dplyr)
library(tibble) # For rownames_to_column
library(knitr)
library(kableExtra)
#---------------------------------------------------------------------

#---------------------------------------------------------------------
# Extract posterior summaries using brms::posterior_summary
#---------------------------------------------------------------------
# Apply posterior_summary to each model, keep results in a list
posterior_summaries_list <- map(
  results_brms_valid$mod_brms,
  ~ posterior_summary(.x, probs = c(0.025, 0.975)) # Get mean, sd, and 95% CrI
)

# --- Define the key parameters to keep in the final table ---
# Verify these names against your actual model output if needed
params_to_keep <- c(
    "b_Intercept",      # Intercept
    "sigma",            # Residual SD for Gaussian
    "shape",            # Shape parameter for Negative Binomial
    "phi",              # Precision for Beta/ZOIB
    "zoi",              # Zero-inflation probability for ZOIB
    "coi",              # One-inflation probability for ZOIB
    "sds_syear_1"       # SD for the smooth term s(year)
)

#---------------------------------------------------------------------
# Process the list into a final data frame
#---------------------------------------------------------------------
tidy_summaries_final <- map2_dfr(
  posterior_summaries_list,
  results_brms_valid$variable,
  ~{
    # Convert matrix output to tibble
    summary_df <- as_tibble(.x, rownames = "Parameter")

    # Filter for desired parameters
    summary_filtered <- summary_df %>%
      filter(Parameter %in% params_to_keep)

    # Add variable name and parameter type classification
    summary_filtered %>%
      mutate(
        variable = .y, # Add variable name
        `Parameter Type` = case_when(
            Parameter == "b_Intercept" ~ "Intercept",
            Parameter %in% c("sigma", "shape", "phi", "zoi", "coi") ~ "Distributional",
            startsWith(Parameter, "sds_") ~ "Smooth SD",
            TRUE ~ "Other"
        ),
        # Rename parameters for better readability
        Parameter = case_when(
            Parameter == "b_Intercept" ~ "(Intercept)",
            Parameter == "sds_syear_1" ~ "sds(s(year))",
            TRUE ~ Parameter
        )
      )
  }
)

# --- Select and order columns for the final table ---
tidy_summaries_final <- tidy_summaries_final %>%
    rename(
        Estimate = Estimate,
        Est.Error = Est.Error,
        `Lower 95% CI` = Q2.5, # Rename quantile column
        `Upper 95% CI` = Q97.5 # Rename quantile column
    ) %>%
    select(
        variable, Parameter, `Parameter Type`,
        Estimate, Est.Error,
        `Lower 95% CI`, `Upper 95% CI`
    ) %>%
    arrange(variable, `Parameter Type`, Parameter)

#---------------------------------------------------------------------
# Create and Display the Formatted Table
#---------------------------------------------------------------------
# Check if the final tibble is empty
if(nrow(tidy_summaries_final) == 0) {
    warning("Final parameter summary table is empty. Check parameter names in 'params_to_keep'.")
} else {
    kable(
      tidy_summaries_final,
      caption = "Posterior summary statistics for key model parameters.",
      digits = 2,
      booktabs = TRUE,
      linesep = ""
    ) %>%
     kableExtra::kable_styling(
        bootstrap_options = c("striped", "hover", "condensed"),
        latex_options = c("striped", "scale_down"),
        full_width = FALSE
        ) %>%
     kableExtra::column_spec(1, bold = TRUE) # Make variable name bold
}
```

The estimated non-linear effect of publication year on each response variable is visualized below (@fig-smooth-plots), showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).

```{r}
#| label: fig-smooth-plots
#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs. The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval. Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."

# Load necessary libraries
library(brms)
library(purrr)
library(dplyr)
library(ggplot2)
library(cowplot)

# Generate conditional smooth plots within map2
conditional_smooths_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    # Assign model and variable name
    model <- .x
    var_name <- .y

    # Generate the plot object list directly
    plot_object_list <- plot(conditional_smooths(model), plot = TRUE)

    # Define the y-axis label based on variable name (link function)
    y_axis_label <- case_when(
        var_name %in% c("N. of authors", "N. of pages") ~ "s(year) Contribution (log scale)",
        var_name == "Recency of references" ~ "s(year) Contribution (logit scale)",
        TRUE ~ "s(year) Contribution" # Default for Gaussian models
    )

    # Extract the first plot, add title, label, and theme
    plot_object_list[[1]] +
      ggtitle(var_name) +
      ylab(y_axis_label) +
      theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
  }
)

# Combine the plots into a grid (adjust ncol as desired)
# The plot_grid call itself will display the combined plot
plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)
